---
title: "Group 3 Bed-Capacity Analysis & Plots"
output:
  html_document: 
   theme: bootstrap
   highlight: monochrome
   toc: yes
   toc_depth: 3
   toc_float: yes
      
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)#Whether to display code inc it's results.
knitr::opts_chunk$set(message = FALSE)#Whether to display messages.
knitr::opts_chunk$set(warning = FALSE)#Whether to display warnings.
knitr::opts_chunk$set(error = FALSE)#Whether to display errors.
knitr::opts_chunk$set(eval = FALSE)#Whether to evaluate the code & incl results.

```




---

## 1. Documentation

<br>


#### 1.1 Purpose of this document.

The purpose of this document is to improve insight into bed capacity across
each of Scotland's Health Boards.  

This document aims to achieve this by analysing open source data made available by 
[ Public Health Scotland (PHS) ](https://www.opendata.nhs.scot/) via
The Scottish Health and Social Care Open Data platform.

This documents focuses on bed capacity across Scotland's Acute Hospital Activity.

This group contains datasets on outpatient activity, inpatient and day case 
activity, the number of NHS beds derived from the [ Acute Hospital Activity and 
NHS Beds Information publication from ISD Scotland ](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-annual/acute-hospital-activity-and-nhs-beds-information-annual-annual-year-ending-31-march-2020/). 

Note that 'Acute' hospital care excludes obstetric, psychiatric and 
long stay care services that are covered in other groups and datasets.

<br>

#### 1.2 Domain Definitions

<br>


| **Bed Definition** | **Description** |
| --- | --- | 
| Available Staffed Beds | This reflects the number of beds that can be used for inpatient or daycase care, multiplied by the number of days in a time period i.e. number of available staffed bed days over the quarter: **Total number of available staffed beds (aasb) over the quarter = Allocated Beds + Borrowed Beds – Lent Beds + Temporary Beds** | 
 Average Available Staffed Beds | This is the average daily number of beds, which are staffed and available for the reception of patients (borrowed and temporary beds are included):**Average number of available staffed beds per day (asb) = aasb / number of days in the quarter**  |
 Allocated Beds | An [ allocated bed](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=A&ID=76&Title=Allocated%20Bed/) is a bed from the specialty bed complement which is resourced for use by the specialty.Beds temporarily out of use remain part of the specialty bed complement but are not available for use, normally for inpatient and day cases. | 
 Borrowed Beds | A [ borrowed bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=B&ID=117&Title=Borrowed%20Bed/) is a bed which is made available to a specialty/significant facility other than the specialty/significant facility to which it is allocated. |
 Lent Beds | A [ lent bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=L&ID=307&Title=Lent%20Bed/) is a bed which is donated temporarily by a specialty/significant facility to another specialty/significant facility. |  
 Temporary | A [temporary bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=T&ID=492&Title=Temporary%20Bed/ ) is a bed erected in a ward, additional to the bed complement of the ward, Note: Temporary beds are used at times of overcrowding and are normally located in the area of accommodation of the specialty involved. |
 Occupied Bed | An occupied bed is an available staffed bed, which is either being used to accommodate an inpatient or reserved for a patient on pass: **Total number of occupied bed days (tobd) = Sum of the number of occupied beds for each day of the quarter. Average number of occupied beds per day (aob) = tobd / number of days in the quarter.** |
 Percentage Occupancy (%) | The percentage occupancy is the percentage of average available staffed beds that were occupied by inpatients during the period: Percentage occupancy = (aob / asb) x 100 |

---

<br>

## 2.  Data 

<br>

#### 2.1 Data Background & Quality 

<br>

Outpatient, inpatient and daycase activity data are collected across NHS Scotland 
and arebased on nationally available information routinely drawn from hospital 
administrative systems across the country. The principal data sources are:

* SMR00 (patient-level outpatient records) - source for outpatients (except return
attendances

* SMR01(inpatients and daycases discharges from non-obstetric and non-psychiatric
specialties) - source for acute inpatients and daycases;

* ISD(S)1 (aggregate hospital activity) - source for bed data returns and return
outpatients. ISD(S)1 is a set of aggregated summary statistics on activity in 
hospitals in NHS Scotland and is derived from monthly and quarterly returns from 
the NHS Boards. 

**Note:  ISD(S)1 is the only source of bed occupancy and bed availability 
data and contains summarised data by NHS Board of Treatment, hospital and 
specialty. ISD(S)1 is also used for return outpatient activity since completeness 
for return outpatients in SMR00 is poor**

<br>

#### 2.2 Data Revision

<br>

NHS Boards can update both their current and historical data monthly. This may 
result in changes in the used data sources.

The data used in this document was last updated	22 February 2022, 
11:30 (UTC+00:00).

<br>

---

## 3. Data Analysis - Beds Occupancy.

<br>

The following code chunks were transferred to the dashboard UI and Server
scripts. The interpretation of the analysis is dependent upon the filters selected 
as UI and is therefore subject to change. Plots can be viewed within dashboard.


```{r}

# The first step in the analysis was to run the library's available on the 
# global.r script for this project.  

# The next step was to load in the bed dataset from the group joined data set
# script. 

bed_data <- read_csv(here("raw_data/nhs_data_joined4.csv")) 

```


```{r, results="hide" }
#The next step is to view the date, using skimr package.

bed_data %>% 
  skimr::skim()

```


> **Plot 1 - Average Bed Occupancy (%) by Hospital**

```{r, echo=TRUE}

plot1 <- bed_data %>%
  select(location_name, percentage_occupancy) %>% 
##UI - Add a filter column her by location name linked to ui, allow multiple selections.
  group_by(location_name) %>% 
  summarise(location_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                            location_total)) %>% 
  mutate(quartiles = cut_number(x = average_bed_occupancy,n = 4, 
                                labels = c("AveOcc in Lowest 25%",
                                           "AveOcc in Lowest 50%",
                                           "AveOcc in Highest 50%", 
                                           "AveOcc in Highest 75%")) )%>% 
  ggplot()+
  aes(x = location_name, y = average_bed_occupancy, fill = quartiles)+
  geom_col()+
  theme_bw() +
  coord_flip()+
  labs(title = "Average Bed Occupancy (%) by Hospital", 
       x = "Location Name", y = "Average Bed Occupancy (%)")+
  scale_fill_discrete(name = "Quartile")

ggplotly(plot1)

```


> **Plot 2 - Average Bed Occupancy by Season **


```{r, echo=TRUE}

plot2 <- bed_data %>%
  select(location_name, percentage_occupancy, season) %>%  
##UI - Add a filter column here by location name linked to ui, allow multiple selections.
  group_by(location_name, season) %>% 
  summarise(location_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                        location_total)) %>% 
  ggplot()+
  aes(x = location_name, y = average_bed_occupancy, shape = season, colour = season)+
  geom_point()+
  theme_bw()+
  coord_flip()+
labs(title = "Average Bed Occupancy (%) by Season", 
     x = "Location Name", y = "Average Bed Occupancy (%)")
 
ggplotly(plot2)

```

> **Plot 3 - Hospital Average Occupancy (%) by Specialty (All) Hospitals**

```{r, echo=TRUE}

plot3 <- bed_data %>%
  select(specialty_name, percentage_occupancy) %>% 
  filter(!specialty_name %in% c("All Specialties", "All Acute")) %>% 
##UI - Add a filter column here by specialty name linked to ui, allow multiple selections.
  filter(specialty_name == "Physiotherapy" | specialty_name == "Infectious Diseases") %>% 
  group_by(specialty_name) %>% 
  summarise(specialty_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                        specialty_total)) %>% 
  arrange(desc(average_bed_occupancy)) %>% 
  ggplot()+
  aes(x = reorder(specialty_name, desc(average_bed_occupancy)), 
      y = average_bed_occupancy, fill = specialty_name)+
  geom_col()+
  theme_bw() +
labs(title = "Average Bed Occupancy (%) by Specialty" , 
     x = "Specialty Name", y = "Average Bed Occupancy (%)")+
theme(legend.position = "none", axis.text.x=element_blank())+
  scale_x_discrete(guide = guide_axis(check.overlap = TRUE))

ggplotly(plot3)

```

> **Plot 4 - Average Bed Occupancy - Selected Hospital by Quarter**

```{r, echo=TRUE}

plot4 <- bed_data %>%
  #Create date variable.
  mutate(year_quarter = yearquarter(quarter), .after = quarter) %>% 
  filter(!is.na(year_quarter)) %>% 
  filter(location_name == "Aberdeen Royal Infirmary" |
         location_name == "St John's Hospital") %>% 
#UI - You can add a filter column here by location name linked to ui, allow multiple selections.
  group_by(year_quarter, location_name) %>% 
  summarise(location_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                        location_total)) %>% 
  ggplot()+
  aes(x = year_quarter, y = average_bed_occupancy, colour = location_name)+
  geom_line()+
  geom_point()+
  scale_alpha_continuous()+
labs(title = "Average Bed Occupancy (%) by, Location & Quarter (2018 Q1 Onwards)")
  

  ggplotly(plot4) 


```


> **Plot 5 - Average Bed Occupancy - Selected Specialty by Quarter**


```{r, echo=TRUE}

plot5 <- bed_data %>%
  #Create date variable.
  mutate(year_quarter = yearquarter(quarter), .after = quarter) %>% 
  filter(!is.na(year_quarter)) %>% 
  filter(specialty_name == "General Medicine" |
         specialty_name == "Accident & Emergency") %>% 
#UI - You can add a filter column here by location name linked to ui, allow multiple selections.
  group_by(year_quarter, specialty_name) %>% 
  summarise(specialty_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                        specialty_total)) %>% 
  ggplot()+
  aes(x = year_quarter, y = average_bed_occupancy, colour = specialty_name)+
  geom_line()+
  geom_point()+
  scale_alpha_continuous()
labs(title = "Average Bed Occupancy (%) by, Specialty & Quarter (2018 Q1 Onwards)")
  

  ggplotly(plot5) 


```


<br>

## 4. Hypothesis Testing

<br>

<br>

#### 4.1 Hypothesis Test 1 - Difference in Average Bed Occupancy (Two Hospitals)

<br>

Perform an appropriate statistical test to determine whether 
there is a statistically significant difference in average bed occupany between 
two selected hospitals. 

<br>

H0: There is no difference in average occupancy between two selected hospitals.

$$ H_0: \mu_{ave\ bed \ occupancy \ (hospital\ 1)} -\mu_{ave \ bed \ occupancy \ (hospital\ 2)} = 0 $$

Ha: There is a difference in average occupancy between two selected hospitals

$$ H_a: \mu_{ave \ bed \ occupancy \ (hospital\ 1)} -\mu_{ave \ bed \ occupancy \ (hospital\ 2)} \neq 0$$

<br>

#### 4.1a Calcuate Observed Statistic 

<br>

```{r, echo=TRUE}
#Step 1 of the bootstrap will be to filter the bed occupancy data by the users selection.
#in the below code i have selected the UI for testing code. 

average_bed_occupancy_twohospitals <- bed_data %>% 
#UI - The below filter would be linked to the UI hospital selection. We can also
  # add a season filter.
  filter(location_name == "Western General Hospital" | 
           location_name == "Belford Hospital") %>%
  filter(season == "Autumn") %>% 
  group_by(location_name) 

observed_ave_occupancy_stat1 <- average_bed_occupancy_twohospitals %>% 
  specify(percentage_occupancy ~ location_name) %>%
  #Order should be dictated by UI location name selection)
  calculate(stat = "diff in means", 
            order = c("Western General Hospital", "Belford Hospital"))
  
observed_ave_occupancy_stat1

```
<br>

#### 4.1b Null Distribution Calculation 

<br>

```{r, echo=TRUE}
# This code creates a null distribution to determine if 
#observed_ave_occupancy_stat1 is significant. 

null_dist1 <- average_bed_occupancy_twohospitals %>% 
#testing relationship between hospital and average bed occupancy.
  specify(percentage_occupancy ~ location_name) %>% 
#the null hypothesis is there is no relationship i.e. they are independent
  hypothesise(null = "independence") %>% 
  generate(reps = 10000, type = "permute") %>% 
#Sample stat is mean of hospitals done in this order.
  calculate(stat = "diff in means", 
            order = c("Western General Hospital", "Belford Hospital"))

head(null_dist1)
```
<br>

#### 4.1c  Probability Value (P-Value)

<br>

```{r, echo=TRUE}
# This calculation generate probability of observing observed stat.
#Selection of one-tailed test.
p_value1 <- null_dist1 %>%
  get_p_value(obs_stat = observed_ave_occupancy_stat1, direction = "both")
  
p_value1

# Extension - Can we add an if function here to return a phrase whether there is
#a significant difference based on probability value greater than significance value
#(alpha)set at 0.05. Other development options could include UI selection of alpha.

```
<br>

#### 4.1d Visualise Average Bed Occupancy Distribution

<br>

```{r, echo=TRUE}
# Let's visualise the distribution with the observed stat. Direction set to 
#both because we are looking for difference. 

null_dist1 %>%
  visualise(bins = 30) +
  shade_p_value(obs_stat = observed_ave_occupancy_stat1, direction = "both") +
  labs(title = "Simulation Based Null Distribution (Diff in Means)", x = "Null Distribution" , y = "Count", 
       caption = paste0("Observed Stat:", round(observed_ave_occupancy_stat1), "\n",
                        "Prob Value: ", p_value1, "\n", "alpha = ", 0.05))
  

```
<br>

#### 4.1e Calculate Confidence Intervals (CI) & Visualisation

<br>

```{r, echo=TRUE}
# This code generates the confidence intervals (upper and lower) for the
#generated null distribution.

bed_ci_95_twohospitals <- null_dist1 %>%
  get_confidence_interval(level = 0.95, type = "percentile")

bed_ci_95_twohospitals

```

```{r, echo=TRUE}
# This code generates the visual of confidence intervals (upper and lower) for the
#generated null distribution.

null_dist1%>%
  visualise(bins = 30) +
  shade_confidence_interval(endpoints = bed_ci_95_twohospitals)+
  labs(title = "Distribution (95% Confidence Interval) ", x = "Null Distribution (CI)" , y = "Count", 
       caption = paste0("CI Lower: ", round(bed_ci_95_twohospitals$lower_ci, digits = 3), "\n",
                        "CI Upper: ", round(bed_ci_95_twohospitals$upper_ci, digits = 3)))
                        
```
<br>

#### 4.2 Hypothesis Test 2 - Difference in Average Bed Occupancy (Hospital Vs Scotland)

<br>
Perform an appropriate statistical test to determine whether 
there is a statistically significant difference in average bed occupancy between 
one hospital and the average bed occupancy for Scotland.

<br>


H0: There is no difference in average occupancy between selected hospital and Scotland.

$$ H_0: \mu_{average\ bed\ occupancy \ (scotland)} -\mu_{ave \ bed \ occupancy \ (hospital)} = 0 $$

Ha: The average occupancy of the selected hospital is greater than average for all of Scotland

$$ H_a: \mu_{ave \ bed \ occupancy \ (scotland)} -\mu_{ave \ bed \ occupancy \ (hospital)} \neq0 $$
<br>

#### 4.1a Calcuate Observed Statistic 

<br>

```{r, echo=TRUE}
#Calculate Average Occupancy for selected hospital.

average_bed_occupancy_onehospital_stat2 <- bed_data %>% 
#UI - The below filter would be linked to the UI hospital selection. 
  filter(location_name == "Western General Hospital") %>% 
  group_by(location_name) %>% 
    summarise(total = n(), 
              average_bed_occupancy_onehospital = sum(percentage_occupancy, 
                                                      na.rm = TRUE)/total)
  

average_bed_occupancy_onehospital_stat2
```
<br>

#### 4.2b Null Distribution Calculation 

<br>
```{r, echo=TRUE}
#Calculate Average Occupancy for All of Scotland Hospitals. 
#Then create null distribution around this mean.

ave_occupancy_scotland <- bed_data %>%
  summarise(total = n(), average_bed_occupancy = round(sum(percentage_occupancy,
                                                           na.rm =TRUE)/ 
                                            total))

ave_occupancy_scotland
```

<br>

```{r, echo=TRUE}
# Bootstrap on calculated Average Bed Occupancy for all of Scotland (No filters).
#Mean used in calculation is from previous code line. 

null_dist2 <- bed_data %>% 
  specify(response = percentage_occupancy) %>% 
  #We are hypothesing that our mean rating for all of Scotlands hospitals will be
  # = to our calculated value (i.e. no difference)
  hypothesize(null = "point", mu = ave_occupancy_scotland$average_bed_occupancy) %>% 
  generate(reps = 10000, type = "bootstrap") %>% 
calculate(stat = "mean")

```
<br>

#### 4.2c  Probability Value (P-Value)

<br>
```{r, echo=TRUE}
# This calculation generate probability of observing observed stat.
#Selection of one-tailed test.
p_value2 <- null_dist2 %>%
  get_p_value(obs_stat = average_bed_occupancy_onehospital_stat2$average_bed_occupancy_onehospital, 
              direction = "both")
  
p_value2

# Extension - Can we add an if function here to return a phrase whether there is
#a significant difference based on probability value greater than significance value
#(alpha)set at 0.05. Other development options could include UI selection of alpha.

```
<br>

#### 4.2d Visualise Average Bed Occupancy Distribution

<br>

```{r, echo=TRUE}
# Lets's visualise the distribution with the observed stat. Direction set to 
#both because we are looking for difference. 

null_dist2 %>%
  visualise(bins = 30) +
  shade_p_value(obs_stat = average_bed_occupancy_onehospital_stat2$average_bed_occupancy_onehospital, direction = "both") +
  labs(title = "Simulation Based Null Distribution (Diff in Means)", 
       x = "Null Distribution" , y = "Count", 
       caption = paste0("Observed Stat:", round(average_bed_occupancy_onehospital_stat2$average_bed_occupancy_onehospital),
                        "\n",
                        "Prob Value: ", p_value2, "\n", "alpha = ", 0.05))

```
<br>

#### 4.2e Calculate Confidence Intervals (CI) & Visualisation

<br>

```{r, echo=TRUE}
# This code generates the confidence intervals (upper and lower) for the
#generated null distribution.

bed_ci_95_onehospital <- null_dist2 %>%
  get_confidence_interval(level = 0.95, type = "percentile")

bed_ci_95_onehospital
```

```{r, echo=TRUE}
# This code generates the visual of confidence intervals (upper and lower) for the
#generated null distribution.

null_dist2%>%
  visualise(bins = 30) +
  shade_confidence_interval(endpoints = bed_ci_95_onehospital)+
  labs(title = "Distribution (95% Confidence Interval) ", x = "Null Distribution (CI)" , y = "Count", 
       caption = paste0("CI Lower: ", round(bed_ci_95_onehospital$lower_ci, digits = 3), "\n",
                        "CI Upper: ", round(bed_ci_95_onehospital$upper_ci, digits = 3)))
                        
```



**End of Analysis Script**
