---
title: "Group 3 Bed-Capacity"
output:
  html_document: 
   theme: journal
   highlight: monochrome
   toc: yes
   toc_depth: 3
   toc_float: yes
   code_folding: hide  
      
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)#Whether to display code inc it's results.
knitr::opts_chunk$set(message = FALSE)#Whether to display messages.
knitr::opts_chunk$set(warning = FALSE)#Whether to display warnings.
knitr::opts_chunk$set(error = FALSE)#Whether to display errors.
knitr::opts_chunk$set(eval = FALSE)#Whether to evaluate the code & incl results.

```


# 1. Documentation

<br>


#### Purpose of this document.

The purpose of this document is to improve insight into bed capacity across
each of Scotland's Health Boards.  

This document aims to achieve this by analysing open source data made available by 
[ Public Health Scptland (PHS) ](https://www.opendata.nhs.scot/) via
The Scottish Health and Social Care Open Data platform.

This documents focuses on bed capacity across Scotland's Acute Hospital Activity.

This group contains datasets on outpatient activity, inpatient and day case 
activity, the number of NHS beds derived from the [ Acute Hospital Activity and 
NHS Beds Information publication from ISD Scotland ](https://publichealthscotland.scot/publications/acute-hospital-activity-and-nhs-beds-information-annual/acute-hospital-activity-and-nhs-beds-information-annual-annual-year-ending-31-march-2020/). 

Note that 'Acute' hospital care excludes obstetric, psychiatric and 
long stay care services that are covered in other groups and datasets.


#### 1a. Domain Knowledge 

The average number of available hospital beds in Scotland has generally been
decreasing over the years. In 2019/20, the average available staffed beds for acute
specialties was 13,156; a small 0.5% increase on last year (2018/19) and a 6%
decrease when compared to five years ago (2014/15).
For 2019/20:
- 9,260 (70%) were for medical specialties
- 3,896 (30%) were for surgical specialties.
The percentage occupancy has remained relatively stable over the years. In
2019/20, the percentage occupancy for acute specialties was 86.7%.

How long a patient stays in hospital will be strongly related to the complexity 
of any operation carried out as well the underlying health condition of the person. 
The average length of stay has been reducing over the years. In 2019/20, 
the average length of stay was 6.1 days compared to 6.9 days in 2010/11. 
Patients admitted as emergencies generally stay longer than elective hospital 
admissions. In 2019/20:

* Elective admissions: the average length of stay was 3.4 days
* Emergency admissions: the average length of stay was 6.7 days

The number of hospital beds has been reducing for many years. This is a result of both
medical advances which have led to shorter stays in hospital for patients including planned
daycase procedures (see Chart 10) alongside a shift to treatment and care in a more
ambulatory setting or in the community.


Background
There are two broad ways in which patients access and make use of acute hospital
services.


* 1. The first is part of a planned or elective pathway of care which is 
normally initiated following a visit to the GP or other healthcare professional, 
and may result in a referral to see a consultant as an **outpatient** for 
specialist advice or diagnosis. This outpatient appointment may then result in 
an onward referral for further tests or admission into hospital for treatment.

<br>

* 2. The second way in which patients make use of hospital services is as a result 
of an emergency referral either by a healthcare professional or directly by the 
patient themselves.This may be via an Accident & Emergency department, directly to
Ambulatory Emergency Care or to an Acute Assessment Unit, where it will be decided 
if the patient needs to be admitted to an **inpatient ward**.


When admitted to hospital, the patient is either treated on a same day basis, 
often referred to as a **daycase**, or as an inpatient, when the patient will
normally spend at least one night in hospital. Some inpatients may be discharged from 
hospital on the same day as their admission.



#### 1b. Definitions Used in this Document  
<br>

* **Admissions;** terminology used to describe hospital activity in the
reported periods. Strictly speaking the activity actually refers to hospital 
discharges in the reported time period rather than admissions. The difference 
between admissions and discharges is of small importance at the level of detail
considered in this document. 

<br>

* **Hospital admissions;** number of **continuous inpatient stays (CIS)** 
in hospital where the patient was admitted.When showing information by CIS, 
the admission type e.g. elective/emergency is determined by the first admitting 
episode. As a result, transfers will generally not appear within the CIS
analysis.When a transfer does appear it is often the result of a patient being 
transferred from another provider unit e.g. outwith Scotland. However, there 
will also be instances where the admission type has been incorrectly coded, 
unfortunately it is not possible to fully ascertain what the correct admission 
type should have been. As a result, you will see that a small proportion of 
transfers do appear within the various tables. 

<br>

* **Episodes of Care and Continuous Inpatient Stays;**
If patient admitted,  care can  be transferred between consultants of different 
specialty's. These separate elements are known as ‘episodes’ of care within each continuous
inpatient stay. For the purpose of this document,hospital admissions are 
defined as the number of continuous inpatient stays in hospital where the patient 
was admitted.


| **Bed Definition** | **Description** |
| --- | --- | 
| Available Staffed Beds | This reflects the number of beds that can be used for inpatient or daycase care, multiplied by the number of days in a time period i.e. number of available staffed bed days over the quarter: **Total number of available staffed beds (aasb) over the quarter = Allocated Beds + Borrowed Beds – Lent Beds + Temporary Beds** | 
 Average Available Staffed Beds | This is the average daily number of beds, which are staffed and available for the reception of patients (borrowed and temporary beds are included):**Average number of available staffed beds per day (asb) = aasb / number of days in the quarter**  |
 Allocated Beds | An [ allocated bed](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=A&ID=76&Title=Allocated%20Bed/) is a bed from the specialty bed complement which is resourced for use by the specialty.Beds temporarily out of use remain part of the specialty bed complement but are not available for use, normally for inpatient and day cases. | 
 Borrowed Beds | A [ borrowed bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=B&ID=117&Title=Borrowed%20Bed/) is a bed which is made available to a specialty/significant facility other than the specialty/significant facility to which it is allocated. |
 Lent Beds | A [ lent bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=L&ID=307&Title=Lent%20Bed/) is a bed which is donated temporarily by a specialty/significant facility to another specialty/significant facility. |  
 Temporary | A [temporary bed ](https://www.ndc.scot.nhs.uk/Dictionary-A-Z/Definitions/index.asp?Search=T&ID=492&Title=Temporary%20Bed/ ) is a bed erected in a ward, additional to the bed complement of the ward, Note: Temporary beds are used at times of overcrowding and are normally located in the area of accommodation of the specialty involved. |
 Occupied Bed | An occupied bed is an available staffed bed, which is either being used to accommodate an inpatient or reserved for a patient on pass: **Total number of occupied bed days (tobd) = Sum of the number of occupied beds for each day of the quarter. Average number of occupied beds per day (aob) = tobd / number of days in the quarter.** |
 Percentage Occupancy (%) | The percentage occupancy is the percentage of average available staffed beds that were occupied by inpatients during the period: Percentage occupancy = (aob / asb) x 100 |
 

# 2.Data 

#### 2a. Data Background & Quality 
Outpatient, inpatient and daycase activity data are collected across NHS Scotland 
and arebased on nationally available information routinely drawn from hospital 
administrative systems across the country. The principal data sources are:

* SMR00 (patient-level outpatient records) - source for outpatients (except return
attendances

* SMR01(inpatients and daycases discharges from non-obstetric and non-psychiatric
specialties) - source for acute inpatients and daycases;

* ISD(S)1 (aggregate hospital activity) - source for bed data returns and return
outpatients. ISD(S)1 is a set of aggregated summary statistics on activity in 
hospitals in NHS Scotland and is derived from monthly and quarterly returns from 
the NHS Boards. 

**Note:  ISD(S)1 is the only source of bed occupancy and bed availability 
data and contains summarised data by NHS Board of Treatment, hospital and 
specialty. ISD(S)1 is also used for return outpatient activity since completeness 
for return outpatients in SMR00 is poor**

#### 2b. Revision
NHS Boards can update both their current and historical data monthly. This may 
result in changes in the used data sources.

The data used in this document was last updated	22 February 2022, 
11:30 (UTC+00:00).

<br>

# 3. Data Analysis - Beds

```{r, results="hide"}
# The first step in the analysis was to run the library's available on the 
# global.r script for this project.  

# The next step was to load in the bed dataset from the group joined data set
# script. 

bed_data <- read_csv(here("raw_data/nhs_data_joined3.csv"))

#Next I loaded in the location look up file. 
location_look_up <- read_csv(here("raw_data/hospital_locations_lookup_file.csv")) %>% 
  clean_names() %>% 
  select(-"id")
```


```{r, results="hide" }
#The next step is to view the date, using skimr package.

bed_data %>% 
  skimr::skim()

location_look_up %>% 
  skimr::skim()
```


```{r, results="hide"}
#Remove columns not required to review bed occupancy & split quarter column.

bed_data_year <- bed_data %>% 
#change variable name to reflect definitions.
  rename("available_staffed_beds" = "all_staffed_beds") %>% 
<<<<<<< Updated upstream
  select(1:20) %>% 
#split quarter variable into quarter & year.
mutate(quarter_year = str_remove(quarter, pattern = "([Q]*[1-4]+$)"), 
       .before = quarter) %>% 
mutate(quarter_year = as.numeric(quarter_year)) %>% 
mutate(quarter = str_remove(quarter, pattern = "(^[0-9]+[0-9]+[0-9]+[0-9]+)"), 
       .after = quarter_year ) %>% 
#round variables
=======
  select(-c(percentage_occupancy_qf:average_daily_number_of_delayed_beds)) %>% 
>>>>>>> Stashed changes
mutate(average_available_staffed_beds = round(average_available_staffed_beds),
       average_occupied_beds = round(average_occupied_beds), 
       percentage_occupancy = round(percentage_occupancy)) %>% 
filter(!is.na(quarter_year)) %>% 
#filter years to 2018 onwards, as agreed with project group.  
filter(quarter_year >= 2018 & quarter %in% c("Q2", "Q3", "Q4"))

```



```{r, results="hide"}
#Upon review of the data it was identified that every second observation, 
# had a statistical qualifier "d" (see qualifier information for definition). 
#Based on the definition, and the intent of this analysis these observations 
#were removed. 

#This was filtered out prior to a join with the look_up file. 

#join location look-up file to bed_data
joined_bed_data <- bed_data_year %>% 
  filter(!location_qf %in% "d") %>% 
  left_join(location_look_up, by = "location") %>% 
<<<<<<< Updated upstream
#End joined table at postcode variable.
  select(1:27)
=======
#End joined table at postcode variable. Note will need to change if joined data
  #set 3 changes. 
  select(-c(address_line:based_on_postcode) )
>>>>>>> Stashed changes

```


> **Plot 1 - Hospital Average Occupancy (%) Compared to other Scottish Hospitals:**


```{r, echo=TRUE}
plot1 <- joined_bed_data %>%
  select(location_name, percentage_occupancy) %>% 
  group_by(location_name) %>% 
  summarise(location_total = n(), 
            average_bed_occupancy = round(sum(percentage_occupancy, na.rm =TRUE)/ 
                                        location_total)) %>% 
  mutate(quartiles = cut_number(x = average_bed_occupancy,n = 4, 
                        labels = c("AveOcc in Highest 25%",
                                   "AveOcc in Highest 50 % ",
                                   "AveOcc in Lowest 50%", 
                                   "AveOcc in Lowest 25%"))
                                           ) %>% 
  ggplot()+
  aes(x = location_name, y = average_bed_occupancy, fill = quartiles)+
  geom_col()+
  theme_bw() +
  coord_flip()+
labs(x = "Location Name", y = "Average Occupancy (%)")

ggplotly(plot1)
       

#Checked NA's prior to removal 8 in total across Dr Grays Hospital & Borders. 
# Should a verification be added here in terms of NA's exceed percentage of 
#location total entries.
```

